---
title: "Tornado Data EDA"
author: "LCM"
format: html
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggthemes)

theme_set(theme_minimal(base_size = 12))
```



## The Data

We obtained the raw data from [here]("https://www.spc.noaa.gov/wcm/data/1950-2022_actual_tornadoes.csv"). It cotains observations about tornado events from 1950 - 2022. The data is downloaded and cleaned in the R script 'get_tornados_data.R'. The cleaned data is saved in the data folder as 'tornado.csv'

Let's load the data:

```{r}
tornados <- read_csv("data/tornados.csv")
```

Check what we read in

```{r}
glimpse(tornados)
```


This data contains 68,693 rows and 27 columns (variables). 

## Exploratory Data Analysis

Initial examination of entire data 

```{r}
skimr::skim(tornados)
```

Note: There are a lot of missing values for 'loss' and a fair number for 'mag' as well. 


```{r}
tornados %>%
  ggplot(aes(x = st)) + 
  geom_bar() + 
  coord_flip()
```


Note: There are 50 states and 3 territories included. 



```{r}
states_df <- map_data("state")
```

Build our map data and filter out AK, HI, PR, and VI.

```{r}
states_df <- map_data("state")
tornados_49 <- tornados %>%
  filter(st != "AK", st != "HI",st!= "PR", st!= "VI")
```



```{r}

tornados_49 %>%
  group_by(st) %>%
  summarise(per_state = n()) %>%
  arrange(desc(per_state)) %>%
  ggplot(aes(x=reorder(st,per_state),y=per_state)) + 
  geom_bar(stat = "identity") + 
  coord_flip()
```

The states from which we see the most tornados:

```{r}
tornados_49 %>%
  group_by(st) %>%
  summarise(per_state = n()) %>%
  arrange(desc(per_state))
```




Filter lut (that is, remove) those observations where the ending lat or long is 0. 

```{r}
tornados_49 <- tornados_49 %>%
  filter(elat > 0 & elon < 0)
```




```{r}
tornados_49 %>%
  ggplot() + 
  geom_polygon(data=states_df,
               mapping = aes(long,lat,group=group),
               fill="white",color="darkgray") + 
  geom_segment(aes(x=slon,y=slat,xend=elon,yend=elat),alpha=0.2)
```




Remove the observations from the West.

```{r}
tornados_49 <- tornados_49 %>%
  filter(slon >= -105)

states_df <- states_df %>%
  filter(long >= -105)
```


```{r}
tornados_49 %>%
  ggplot() + 
  geom_polygon(data=states_df,
               mapping = aes(long,lat,group=group),
               fill="white",color="darkgray") + 
  geom_segment(aes(x=slon,y=slat,xend=elon,yend=elat),alpha=0.5,color="darkgreen")
```

Frequency of tornados by year.

```{r}
tornados_49 %>%
  ggplot(aes(x= yr)) + 
  geom_bar()
```



Tornado frequency by month.

```{r}
tornados_49 %>%
  ggplot(aes(x=mo)) + 
  geom_bar()
```



```{r}
tornados_49 %>%
  ggplot() + 
  geom_polygon(data=states_df,
               mapping = aes(long,lat,group=group),
               fill="white",color="darkgray") + 
  geom_point(aes(x=slon,y=slat,color=mag))

```




















